---
title: "Bespoke Bayesian Model for High Throughput Biochemical Assays"
author: "Anders Ellegaard"
date: '2021-12-30'
bibliography: references.bib
csl: citation_style.csl
link-citations: yes
linkcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment=NA)
```

I am on a quest to improve the model fitting I do on biochemical assays. For some time, I have had this feeling that I should be able to extract more information from the data gathered in biochemical assays, in particular assays with a high throughput.

In [a previous study]({{< relref "/project/bespoke-biochem-one" >}})

```{r utilities, message=FALSE}
library(ggplot2)
library(magrittr)

options(mc.cores = parallel::detectCores())

colour <- list(
  orange_dark = "#fb8500",
  orange_light = "#ffb703",
  blue_dark = "#023047",
  azure = "#219ebc",
  blue_light = "#8ecae6"
)

set.seed(4444)
```

# Generative Process

```{r generative_model}
hill_function <- function(log_conc, bottom, top, log_IC50, nH) {
  top + (bottom - top)/(1 + 10^((log_IC50 - log_conc)*nH))
}

assay_response <- function(log_conc, bottom, top, log_IC50, nH, sigma) {
  noise <- rnorm(length(log_conc), 0, sigma)
  hill_function(log_conc, bottom, top, log_IC50, nH) + noise
}

n_compounds <- 100
assay_window <- seq(-9, -3, length.out = 6)

true_parameters <- tibble::tibble(
  compound = seq(1, n_compounds),
  bottom = 1 - rlnorm(n_compounds, -0.25, 0.125),
  log_IC50 = rnorm(n_compounds, -7, 0.5),
  top = 0.99,
  nH = 1.01,
  sigma = 0.15
)

true_curves <- purrr::pmap(
  true_parameters,
  ~ geom_function(
    fun = hill_function,
    args = list(
      top = ..4,
      bottom = ..2,
      nH = ..5,
      log_IC50 = ..3
    ),
    colour = colour$blue_dark,
    alpha = 0.5
  )
)

p <- ggplot() +
  xlim(-9, -3) +
  theme_minimal() +
  labs(x = "Ligand concentration [M]" , y = "True tissue response")

Reduce(`+`, true_curves[1:10], init = p)
```


```{r generative_model2}
screening_experiment <- function(log_conc,
                                 n_compounds = NULL,
                                 bottom_mean = NULL,
                                 bottom_sd = NULL,
                                 top_mean = NULL,
                                 top_sd = NULL,
                                 log_IC50_mean = NULL,
                                 log_IC50_sd = NULL,
                                 nH_meanlog = NULL,
                                 nH_sdlog = NULL,
                                 sigma_rate = NULL,
                                 parameters = NULL) {
  if (is.null(parameters)) {
    parameters <- tibble::tibble(
      compound = seq(1, n_compounds),
      bottom = rnorm(n_compounds, bottom_mean, bottom_sd),
      log_IC50 = rnorm(n_compounds, log_IC50_mean, log_IC50_sd),
      top = rnorm(1, top_mean, top_sd),
      nH = rlnorm(1, nH_meanlog, nH_sdlog),
      sigma = rexp(1, sigma_rate)
    )
  }
  parameters %>% 
    tidyr::expand_grid(log_conc = log_conc) %>%
    dplyr::mutate(
      response = assay_response(log_conc, bottom, top, log_IC50, nH, sigma)
    )
}

observations <- screening_experiment(log_conc = assay_window, parameters = true_parameters)
```

# Model

In our screening assay, we will consider $M$ compounds $j = 1, ..., M$. For each compound, we measure an assay response, $y_{ij}$, for a number of ligand concentrations $[A_{ij}]$. We also know that the assay response averages to the tissue response, $\mu_ij$, but that observations are noisy:

$$y_{ij} \sim {\sf Normal}(\mu_{ij}, \sigma)$$

Note that the noise parameter, $\sigma$, is the same for all $M$ compounds.

The tissue response is a deterministic function of four kinetic parameters, as described by the Hill equation:

$$\mu_{ij} = top - \frac{bottom_j - top}{1 + 10^{(\log_{10}(IC_{50,j}) - \log_{10}([A_{ij}]))^{n_H}}}$$



# Priors

## Prior Predictive Simulation

```{r}


priors <- list(
  bottom_mean <- 0.25,
  bottom_sd <- 0.25,
  top_mean <- 1,
  top_sd <- 0.025,
  log_IC50_mean <- -6,
  log_IC50_sd <- 0.7,
  nH_meanlog <- 0,
  nH_sdlog <- 0.01,
  sigma_rate <- 10
)


rlang::exec(
  screening_experiment,
  log_conc = assay_window,
  n_compounds = 50,
  !!!priors
) %>% 
  ggplot(aes(x = log_conc, y = response, group = compound)) +
    geom_line()


```

```{r}
data <- list(
  N = nrow(experiment$observations),
  M = max(experiment$observations$curve),
  curve_ind = experiment$observations$curve,
  log_conc = experiment$observations$log_conc,
  y = experiment$observations$response
)

post <- rstan::stan(
  file = "hill_equation_screening.stan",
  data = data,
  chains = 4,
  cores = 4,
  seed = 4444
)

posterior_samples <- rstan::extract(post) %>% tibble::as_tibble()
```

